Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-24 manuscript.rmd
Included some of Laura and Rosies suggestions in the results

\
\     We used a locally parameterized and calibrated 1-D hydrological water balance model to obtain daily dynamics of soil water potential by depth in the whole-soil column and then applied a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water potential best explained observed dynamics of species-specific growth for large trees of `r length(erd.sp)` dominant species (Fig. \@ref(fig:schematic)). This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by leveraging existing data for eight species and developing trait-based proxies for the rest. We evaluated ERD against existing dueterium to hydrogen content data for a subset of six tree species as an independent observation of ERD, given the corresponding gradient in isotopic composition of water in the soil profile. We tested relationship of ERD with above-ground hydraulic traits for a subset of eight species with both datasets. We finally analysed species' mortality rates across the censuses spanning 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought. Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability. Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers. To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

@@

The ELM-FATES simulations with the chosen best-fit parameter ensembles (Table \@ref(tab:param)) successfully captured the dynamics of soil moisture by depth (Fig. \@ref(fig:hydro-calib) A & Fig. \@ref(fig:hydro-calib)B), including the observations from the horizontal TDR probes that were not included in the calibration. These simulations also captured the dynamics in stream discharge (Fig. \@ref(fig:hydro-calib)C) and to a moderate extent the dynamics in evapotranspiration (Fig. \@ref(fig:hydro-calib)D). The reduction in parameter range in the best-fit ensembles compared to the global ranges tested showed that the above calibration was sensitive to the following parameters: the Ball-Berry stomatal slope parameter, the ELM rooting distribution parameter that regulates the depth of the rooting profile, soil hydraulic conductivity profile especially at depth, and our adjustment factor to soil hydraulic conductivity to account for macroporosity and direct flow paths (Table \@ref(tab:param)).

@@

\   Soil Water Potential dynamics ($\Psi_{\textrm{soil},z}$) by depth $z$ (from 0.1 through 13 m) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig. \@ref(fig:swp-droughts)). Notably, soil droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme soil drought years each, depending on depth location (Fig. S\@ref(fig:swp-droughts)). Across the simulation period (1990-2015) $\Psi_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m (Fig. S\@ref(fig:swp-all-depths)).

@@

Our analyses revealed that co-occurring large canopy tree species with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety (Fig. \@ref(fig:erd-hyd)). Species with deeper ERD showed higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species over the 35-yr record (1981-2015) (Fig. S\@ref(fig:erd-mort-census), especially the evergreen species (Fig. \@ref(fig:erd-mort-census-evg)). Analysis of time spent below ($\Psi$~crit~) in the soil zone matching ERD $\Psi_{\textrm{soil},z=\textrm{ERD}}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD (Fig. \@ref(fig:erd-time); Fig. S\@ref(fig:erd-time-evg)). We thus demonstrate, for the first time, a link between realized drought experience, hydraulic architecture and long-term mortality outcomes for co-occurring adult tree species.

@@

Our study brings attention to the need for assessing drought-sensitivity in terms of species realized drought risk by accounting for the actual hydrological drought and trees' deep water access, beyond the universal drought indices and trees' above-ground hydraulic vulnerability. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins, that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through periods of significant droughts (Figs. \@ref(fig:erd-mort-census-evg),  S\@ref(fig:erd-mort-census), \@ref(fig:swp-droughts)). These results are consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016]. @Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi$~88,stem~) and had narrower safety-margins ($\Psi$~min~ - $\Psi$~88,stem~). @Nardini:2015 found that a species sourcing deep water in a Karst system showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi$~88,stem~ and narrower safety margin compared to two species sourcing shallow-water.

\
\   Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018 in which species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, extreme drought. Their hydrological model revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity. In contrast, our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa). The difference in our results are likely attributable to site differences in the precipitation regime. In their dry forest site (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016), and thus at BCI deep soil layers are effectively recharged annually (Fig. S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018). And the kind of droughts that buffered the deep rooted species in our study were recurrent sub-annual droughts that extended the dry season into the wet season. The variation in mortality explained by ERD in our study is upto 40%; other causes may explain observed mortality [@McDowell.2018].

@@

\
\   We benefited from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for a few species ($n$ = `r length(sp.exp.param$sp)`), a 'hard' trait. This allowed us to scale-up the leaf vulnerability curves at the community scale ($n$ = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters). WSG alone explained `r round(b.wsg.r2*100,0)`% of the variance in $B$ ($p$ `r  b.wsg.p`). We find that leaf conductance declined more rapidly with declining $\Psi$~leaf~ for species with higher values of WSG and LMA (Fig. S\@ref(fig:obs-plc-curves)). Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but supportive and counter examples both exist [@McCulloh.2011; @Hoffmann.2011].

\
\   Another important piece of information for understanding effective rooting depth is stable water isotopes of xylem water and below-ground sources, in the vadose zone and water table. Although the interpretation of these data is still under research [@Deurwaerder.2020], they are critical for model calibration or validation. The model in @Chitra-Tarak.2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the saprolite found in their dry forest study site (Jean Riotte, pers. comm.). The model of below-ground water-availability in this study (ELM-FATES) is far more advanced compared to that used by @Chitra-Tarak.2018 and was calibrated with data on long-term soil moisture dynamics at a point location (0-0.15m) and plot-wide measurements by depth (0.15, 0.4 and 1 m), as well as, evapotranspiration and stream discharge. @Chitra-Tarak.2018 was calibrated only on stream discharge and water table levels. We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers and the spatial variability within the plot. As in @Chitra-Tarak.2018, we also interpret our ERD estimates as revealing relative differences among species' effective rooting depths rather than in absolute terms.

### Future directions for model development

The trade-offs that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests. The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high $K_\textrm{max,stem}$) and cavitation-resistance (low $\Psi$~50,stem~ or $\Psi$~88,stem~) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016]. We report such a trade-off ($n$ = 26 species) at our study site stem level. We also found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown). The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.

\
\   Our model estimated ERD as a single value, unlike the dynamic ERD modeled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, using corresponding calibration data on isotopically determined water-sourcing depths, as species are expected to shift to deeper soil water sources as shallow soils dry out [@Meinzer:1999]. Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decouples climate events and demographic outcomes. Future studies could better constrain the model with higher resolution growth data such as those from dendrometer bands. This is especially important for modeling seasonally dynamic ERD. However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also changes due to hydrostatic stem flexing (xylem, phloem and bark hydration), the importance of which increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015]. Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018]. Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data and evaluation data ($\mathrm{\delta}$^2^H, Fig. S\@ref(fig:other-erd-mods)). We interpret this result as that VPD and PLC curves may be adequate to explain growth, but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf fall data and assumption of leaf presence during the wet season, and omits inter-annual variation. Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich tropical forests [@Park.2019]. In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees of the same size class, an issue we minimized by limiting ERD estimates to canopy trees. To estimate topographic variation in ERD within or across species future studies may use soil moisture dynamics from a distributed hydrological model [e.g. @Schwantes:2018]. We note that data availability on soil water retention curve ($\Psi$~soil~ vs. VWC), hydraulic conductivity (K~soil~) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model. We recommend wide-spread data collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments. Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models of ERD (this study and @Chitra-Tarak.2018). Although data needs for parameterizing such models [@Christoffersen:2016kd2d; @Sperry.1998] could be greater (e.g. retention curves for roots), which limits immediate community-wide application, such models hold greater promise to improve our understanding of plant physiology [e.g. see @Johnson:2018fad].

@@

Establishing relationships between plant's environment, traits and demographic outcomes is imperative for developing a predictive plant ecology, but trees realized water environments and water-access are rarely studied. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest. The role of deep-water access that we find in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of tropical forest dynamics under current and future climate. Our community-scale framework for modeling effective rooting depths or leaf vulnerability curves indicate the possibilities in expanding the use of critical, hard-earned observations in species-rich forests towards community-scale generalizations.

>>>>

\
\     We used a locally parameterized and calibrated 1-D hydrological water balance model to obtain daily dynamics of soil water potential by depth in the whole-soil column and then applied a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water potential best explained observed dynamics of species-specific growth for large trees of `r length(erd.sp)` dominant species (Fig. \@ref(fig:schematic)). This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by leveraging existing data for eight species and developing trait-based proxies for the rest. We evaluated ERD against existing dueterium to hydrogen content data for a subset of six tree species as an independent observation of ERD, given the corresponding gradient in isotopic composition of water in the soil profile. We tested relationship of ERD with above-ground hydraulic traits,namely, xylem vulnerability to embolism, branch hydraulic safety margins, turgor loss points and maximum stem-specific hydraulic conductivity, for a subset of eight species with both datasets. Finally, ee analysed species' mortality rates across the censuses spanning 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought. Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability. Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers. To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

@@

The 100 ELM-FATES simulations with the chosen best-fit parameter sets (Table \@ref(tab:param)) successfully captured the dynamics of soil moisture by depth (Fig. \@ref(fig:hydro-calib) A & Fig. \@ref(fig:hydro-calib)B), including the observations from the horizontal TDR probes that were not included in the calibration. These simulations also captured the dynamics in stream discharge (Fig. \@ref(fig:hydro-calib)C) and in evapotranspiration (Fig. \@ref(fig:hydro-calib)D), slightly underestimating peak discharge in wet years and slightly overestimating peak ET. The reduction in parameter range in the best-fit ensembles compared to the global ranges tested showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter (that regulates the depth of the rooting profile), soil hydraulic conductivity profile, especially at depth, and the adjustment factor that varies soil hydraulic conductivity to account for macroporosity and direct flow paths (Table \@ref(tab:param)).

@@

\   Soil Water Potential dynamics ($\Psi_{\textrm{soil},z}$) by depth $z$ (from 0.1 through 13 m) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig. \@ref(fig:swp-droughts)). Notably, soil droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme soil drought years each, depending on depth location (Fig. S\@ref(fig:swp-droughts)). Across the simulation period (1990-2015) $\Psi_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m in the simulations (Fig. S\@ref(fig:swp-all-depths)).

@@

Our novel approach for estimating ERD using 25 years of growth, species-specific leaf vulnerability curves, modeled soil moisture through the profile, and observed VPD was consistent with estimates using xylem water isotopes. Our analyses revealed that co-occurring large canopy tree species with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety (Fig. \@ref(fig:erd-hyd)). Species with deeper ERD showed higher maximum stem hydraulic conductance, and greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species over the 35-yr record (1981-2015) (Fig. S\@ref(fig:erd-mort-census), especially amongst the evergreen species (Fig. \@ref(fig:erd-mort-census-evg)). Analysis of time spent below ($\Psi$~crit~) in the soil zone matching ERD $\Psi_{\textrm{soil},z=\textrm{ERD}}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD (Fig. \@ref(fig:erd-time); Fig. S\@ref(fig:erd-time-evg)). We thus demonstrate, for the first time, a link between realized drought experience, hydraulic architecture and long-term mortality outcomes for co-occurring large tree species.

@@

Our study brings attention to the need for assessing drought sensitivity in terms of species realized drought exposure by accounting for the actual hydrological drought and trees' deep water access, beyond the universal drought indices and trees' above-ground hydraulic vulnerability. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins, that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through periods of extreme droughts (Figs. \@ref(fig:erd-mort-census-evg), S\@ref(fig:erd-mort-census), \@ref(fig:swp-droughts)). Deeper ERD species on average had more hydraulically efficient, but risky, suite of traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk. By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\Psi_{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$. Our results are consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016]. @Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi$~88,stem~) and had narrower safety-margins ($\Psi$~min~ - $\Psi$~88,stem~). @Nardini:2015 found that a species sourcing deep water in a Karst system showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi$~88,stem~ and narrower safety margin compared to two species sourcing shallow-water.

\
\   Our result--of lower mortality in deep rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (CT2018) in which species identified by their inverse ERD model as effectively deep rooted died *more* in a rare, prolonged drought. The hydrological model in that study revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity. In contrast, our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa). The difference in the results shown here are likely attributable to site differences in the precipitation regime. In the dry forest site in CT2018 (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016), and thus at BCI deep soil layers are effectively recharged annually (Fig. S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018). The kind of droughts that buffered the deep rooted species in our study were recurrent sub-annual droughts that extended the dry season into the wet season. The variation in mortality explained by ERD in our study is upto 40%, and other causes may explain observed mortality [@McDowell.2018].

@@

\
\   We found a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for a few species ($n$ = `r length(sp.exp.param$sp)`), a 'hard' trait. This allowed us to scale-up the leaf vulnerability curves at the community scale ($n$ = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters). WSG alone explained `r round(b.wsg.r2*100,0)`% of the variance in $B$ ($p$ `r  b.wsg.p`). We find that leaf conductance declined more rapidly with declining $\Psi$~leaf~ for species with higher values of WSG and LMA (Fig. S\@ref(fig:obs-plc-curves)). Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but supportive and counter examples both exist [@McCulloh.2011; @Hoffmann.2011].

\
\   Another important piece of information for understanding effective rooting depth is stable water isotopes of xylem water and below-ground sources, in the vadose zone and water table. Although the interpretation of these data is still under research [@Deurwaerder.2020], they are critical for model calibration or validation. The model in @Chitra-Tarak.2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the saprolite found in their dry forest study site (Jean Riotte, pers. comm.). The model of below-ground water-availability in this study (ELM-FATES) differs from the conceptual water balance model used by CT2018 and was calibrated with data on long-term soil moisture dynamics at a point location (0-0.15m) and plot-wide measurements by depth (0.15, 0.4 and 1 m), as well as, evapotranspiration and stream discharge. The model in CT2018 was calibrated only on stream discharge and water table levels. We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers and the spatial variability within the plot. As in @Chitra-Tarak.2018, we also interpret our ERD estimates as revealing relative differences among species' effective rooting depths rather than in absolute terms.

### Future directions

The relationships that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests, which intrinsically rely on the parameterization of contrasting life history strategies [@Schieter:2012] and the simulation of competition between those strategies. The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high $K_\textrm{max,stem}$) and cavitation-resistance (low $\Psi$~50,stem~ or $\Psi$~88,stem~) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016]. We report such a trade-off ($n$ = 26 species) at our study site stem level. We also found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown). This offers the promise of a greater ability to parameterie the 'hard' hydraulic traits with the abundant soft trait data, thus allowing much better representation of forest hydrodynamics. The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.

\
\   Our model estimated ERD as a single value, unlike the temporally dynamic ERD modeled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, using corresponding calibration data on isotopically determined water-sourcing depths, as species are expected to shift to deeper soil water sources as shallow soils dry out [@Meinzer:1999]. Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decouples climate events and demographic outcomes. Future studies could better constrain the ERD model with higher resolution growth data such as those from dendrometer bands. This is especially important for modeling seasonally dynamic ERD. However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also changes due to hydrostatic stem flexing (xylem, phloem and bark hydration), the importance of which increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015]. Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018]. Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data and evaluation data ($\mathrm{\delta}$^2^H, Fig. S\@ref(fig:other-erd-mods)). We interpret this result as that VPD and PLC curves may be adequate to explain growth, but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf fall data and assumption of leaf presence during the wet season, and omits inter-annual variation. Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich tropical forests [@Park.2019]. In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees of the same size class, an issue we minimized by limiting ERD estimates to canopy trees. To estimate topographic variation in ERD within or across species future studies may use soil moisture dynamics from a distributed hydrological model [e.g. @Schwantes:2018]. We note that data availability on soil water retention curve ($\Psi$~soil~ vs. VWC), hydraulic conductivity (K~soil~) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model. We recommend wide-spread and coordinated collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments. Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) [@Christoffersen:2016kd2d; @Sperry.1998] would be a step change from the current statistical models of ERD (this study and @Chitra-Tarak.2018). Although data needs for parameterizing such models [@Christoffersen:2016kd2d; @Sperry.1998] could be greater (e.g. retention curves for roots), which limits immediate community-wide application, such models hold greater promise to improve our understanding of plant physiology [e.g. see @Johnson:2018fad].

@@

Establishing relationships between environment, traits and demographic outcomes of plants is imperative for developing a predictive plant ecology. Trees realized water environments and water-access are nonetheless rarely studied. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest. We report here that deep-water access plays a role in alleviating mortality of hydrualically vulnerable trees. This has important implications for our predictive understanding of tropical forest dynamics under current and future climate, and suggests that greater emphasis should be given to predictive modeling of tree root allocation and distribution (e.g. @Wijk.2001) as well as trees' hydrological environments. Our community-scale framework for modeling effective rooting depths and leaf vulnerability curves indicate the possibilities in expanding the use of these critical, hard-earned observations in species-rich forests towards community-scale generalizations.
